### 2020-07-03 msr.tex
editing

\glspl{MSR} are one of six advanced reactor designs shortlisted by
the \gls{GIF} in 2001 for promising significant advances in safety,
sustainability, efficiency, and cost over existing designs in operation
today. This has attracted significant attention and resources towards
\gls{MSR} research, most noticeable by the number of start-up companies that
have emerged in recent years touting various \gls{MSR} designs. This chapter
provides a brief history of \glspl{MSR}, followed by the distinctive features
that earned the concept the label of being a Generation IV reactor. Lastly,
we present the reference specifications of the \gls{MSFR} concept studied in
this work.

@@

The first \gls{MSR}, named the \gls{ARE}, dates back to the 1940s,
as part of the US Aircraft Nuclear Propulsion program
\cite{rosenthal_molten-salt_1970}; the molten salt concept was considered due
to the stability of molten salts at high temperatures and neutron radiation.
The 2.5 MW$_{\text{th}}$ reactor was built at \gls{ORNL}, where it achieved
criticality on November 1954 and generated 100 MWh over nine days
\cite{rosenthal_molten-salt_1970}. The fuel
consisted of enriched uranium in a molten salt mixture of NaF, ZrF$_4$, and
UF$_4$, and was moderated by blocks of beryllium oxide. The project ultimately
never came to fruition as the development of intercontinental ballistic
missiles effectively eliminated the need for long-range nuclear-powered
bomber aircraft.

However, the successful demonstration of the \gls{ARE} spurred further
research into adapting \glspl{MSR} for civilian power generation
\cite{rosenthal_molten-salt_1970}. One of the key findings from the
research was that breeding $^{233}$U from $^{232}$Th gave better performance
than breeding $^{239}$Pu from $^{238}$U in thermal-spectrum reactors.
Ultimately, these efforts culminated in the design, construction, and
successful operation of the \gls{MSRE}, a graphite moderated thermal
\gls{MSR}. Although the experiment did not include breeding, scientists at
\gls{ORNL} obtained a wealth of other experimental data and new insights from
the study of this reactor. The \gls{MSRE} had a
graphite-moderated design with a LiF-BeF$_2$-ZrF$_4$-UF$_4$ fuel salt mixture,
initially rated at 10 MW$_{\text{th}}$ but later restricted to 8
MW$_{\text{th}}$ due to a miscalculation of heat transfer capabilities
\cite{haubenreich_experience_1970}.

Design of the \gls{MSRE} commenced in the summer of 1960, with construction
starting in early 1962 \cite{haubenreich_experience_1970}. The reactor
achieved zero-power criticality in June 1965, and 30 days of continuous
operation at full power in December 1966. The reactor operated at full power
for the most of the following 15 months, during which the researchers carried
out various experiments. Soon after shutdown, the $^{235}$U fuel was replaced
with $^{233}$U and in January 1969, the \gls{MSRE} became the first reactor to
run on $^{233}$U fuel. Further experiments were run, including xenon
stripping, fission product deposition, tritium behavior, and plutonium
addition studies, before the \gls{MSRE} was permanently shut down to conserve
remaining funding for other related activities \cite{macpherson_molten_1985}.

Building on the \gls{MSRE}'s hugely successful run, \gls{ORNL} proposed a
new program for the construction and operation of a demonstration reactor
based on the \gls{MSBR} concept that they had developed
\cite{macpherson_molten_1985}. The \gls{MSBR} is a thermal-spectrum, single
fluid reactor with fertile $^{232}$Th isotopes mixed directly into the FLiBe
molten salt for $^{233}$U breeding \cite{gehin_liquid_2016}. Like the
\gls{MSRE}, the \gls{MSBR} relies on continuous online reprocessing to add
fertile material and remove \gls{FP} neutron poisons. As a breeder
reactor, the doubling time (the minimum amount of time required to produce
enough fissile material to start up another \gls{MSBR}) was estimated to be
approximately 22 years. However, \gls{ORNL} failed to secure funding on two
separate occasions in 1972 and 1974; they lost out to the competing
\gls{LMFBR} program which had a head start and wider political and technical
support. Nevertheless, from a technical perspective, two independent
technology evaluation and design studies of the \gls{MSR} had ``reported
favorably on the promise of the system'' \cite{macpherson_molten_1985}.

In spite of this setback, research into \glspl{MSR} continued through the late
1970s. In 1980, \gls{ORNL} published a report describing a new \gls{MSR}
concept, called the \gls{DMSR} \cite{gehin_liquid_2016}. This design was
developed in response to the fuel reprocessing restrictions introduced by
President Ford in 1976; the \gls{DMSR} is designed to operate as a
once-through converter system without fuel reprocessing. While it is largely
fueled by 19.75 \% \gls{HALEU}, the initial core loading includes thorium to
boost its conversion ratio throughout its lifetime. It has a continuous
online feed consisting of \gls{HALEU} to maintain criticality, and denatured
$^{235}$U to keep uranium enrichment levels below nuclear non-proliferation
policy thresholds. The design also includes a gas sparging system for removing
gaseous \glspl{FP}, while noble metals naturally plate out onto the walls of
the coolant loop. A significant drawback in the \gls{MSBR} design is the
extensive neutron damage in the graphite moderator that necessitated frequent
replacement (every four years) throughout its operational lifetime. The
\gls{DMSR} avoids this issue by having a lower power density while maintaining
the overall power output of 2250 MW$_{\text{th}}$. As a result, the graphite
moderator was projected to last for the entirety of the \gls{DMSR}'s design
lifetime.

There was a concurrent program at the UK Atomic Energy Authority for the
development of a 2500 MW$_{\text{e}}$ lead-cooled
\gls{MCFR} concept \cite{smith_assessment_1974}. It is a dual fluid system,
with separate loops for the fuel salt and the blanket salt. The blanket is a
1 m-wide tank surrounding the core. The relatively harder neutron spectrum
arising from the absence of moderators and the choice of chloride over
fluoride salt favors $^{239}$Pu breeding over the thorium cycle. Some
experiments were performed to study molten salt chemistry but no reactor
prototypes were built. The UK program was eventually shut down just like its
US counterpart partly due to the successful demonstration of the Prototype
Fast Reactor which had achieved criticality in 1974.

Following a lull lasting through the late 20th century, \gls{MSR} research
picked up pace due to renewed interest initiated by the \gls{GIF} in 2001.
Today, there are numerous \gls{MSR} concepts under active development led by
various national and commercial bodies. Many differences exist between
different concepts and the only common denominator is the use of nuclear fuel
dissolved in molten salt. Table * is a non-exhaustive list of \glspl{MSR}
under development.

@@

As mentioned in the introduction section, the most significant difference
between \glspl{MSR} and other reactor concepts is the liquid fuel in
\glspl{MSR}; fissile and/or fertile material is dissolved in high temperature,
commonly eutectic mixtures of molten salts. Most \gls{MSR} designs are
circulating-fuel reactors. The primary coolant loop containing the fuel salt
transfers heat through a heat exchanger to the clean, secondary/intermediate
loop. The liquid fuel form allows for continuous online fuel reprocessing,
and the removal of gaseous \glspl{FP} via a gas sparging system.

The flexibility of \glspl{MSR} is best illustrated by the various designs
under development today. Graphite-moderated thermal-spectrum \glspl{MSR} are
typically straightforward \gls{LEU} burners, or $^{232}$Th/$^{233}$U
iso-breeders/breeders, while epithermal- and fast-spectrum \glspl{MSR} have
the additional options of operating as \gls{TRU} fuel burners or
$^{238}$U/$^{239}$Pu breeders. Breeder designs can be further categorized into
one- or two-fluid designs; two-fluid designs feature separate blanket molten
salt mixtures that contain higher proportions of fertile material than the
fuel salt mixture.

@@

\glspl{MSR} are generally deemed to be safer than \glspl{LWR} due to their
reliance on natural physical phenomena for passive safety.
The most immediate and significant safety characteristic is the strong
negative temperature reactivity feedback of the fuel salt. This is due to
greater temperature-induced volumetric expansion in liquid fuel than solid
fuel. Combined with the Doppler broadening of resonance capture cross sections
present in both fuel forms, we expect to observe a smaller temperature
increase following an unprotected reactivity insertion. The overall
temperature reactivity coefficient varies widely between different \gls{MSR}
designs due to other structures, such as moderators and reflectors, present in
the core. In particular, graphite moderators tend to have slightly positive
temperature reactivity coefficients. This was notably observed in the
\gls{MSBR} concept, but the total temperature reactivity coefficient was still
relatively large and negative \cite{rykhlevskii_modeling_2019}. This
phenomenon provides a large degree of control and stability as it is always
present in an \gls{MSR} regardless of the operating conditions.

Continuous online fuel reprocessing allows operators to maintain low excess
reactivity inventories in the core as additional fuel can be added on an ad
hoc basis. Reprocessing and gas sparging systems help reduce fissile
requirements by continuously removing neutron poisons. These factors, in
addition to the strong negative temperature
reactivity coefficient, diminishes the likelihood and severity of unprotected
criticality accidents in \glspl{MSR} \cite{elsheikh_safety_2013}. In the
unlikely situation where an \gls{MSR} encounters a severe runaway reaction,
we can rely on another passive safety feature called freeze plugs. A freeze
plug is a plug of solidified salt at the bottom of the core actively cooled by
fans or other cooling systems to keep its temperature just below the freezing
point of the salt \cite{aji_experimental_2020}. When temperatures in the core
exceed a certain threshold during a dangerous transient, the freeze plug melts
and the molten salt in the core drains into a containment tank designed to
keep the salt in a subcritical configuration. This is especially easy to
achieve for
thermal-spectrum \glspl{MSR} as the absence of moderators in the containment
tank would automatically bring the multiplication factor down below unity
\cite{elsheikh_safety_2013}. \glspl{MSR} also typically have large
margin-to-boiling under nominal operating conditions so that fuel salt boiling
does not occur. Furthermore, the reactor vessel is consequently subject to
much lower stresses as \glspl{MSR} operate at near-atmospheric pressure
levels. Thus, the probability of pipe ruptures due to high pressure is low.

\glspl{MSR} may be more resilient to pump failure accidents as natural
circulation could passively sustain enough heat transfer through the heat
exchanger to remove any remaining fission and decay heat after the failure. If
natural circulation proves insufficient, the aforementioned freeze plug can
drain the salt out of the core. Decay heat in \glspl{MSR} with online
reprocessing is typically lower than that in \glspl{LWR} due to the
continuous removal of \glspl{FP}. For example, the decay heat in an \gls{MSFR}
after reaching equilibrium salt composition is approximately 3.5\% compared
to 6\% in \glspl{LWR} \cite{brovchenko_design-related_2013}.

However, there are also some safety disadvantages associated with \glspl{MSR}.
Firstly, \glspl{MSR} have smaller fractions of \glspl{DNP} in the active core
region as some of them decay in the external loop regions; this complicates
reactor control and may result in faster transients due to the decrease in
average neutron lifetime. Secondly, structural materials in the core must be
resistant to salt corrosion under high temperatures and neutron irradiation
over the lifetime of the reactor. To a lesser extent, pipes should also be
corrosion resistant to prevent pipe ruptures. Lastly, fuel salt overcooling is
a potentially dangerous accident scenario as the strong temperature
coefficient can effectively cause a large reactivity insertion. Overcooling
may also cause salt to freeze and restrict flow, thereby causing a complete
loss of flow accident.

\subsection{Other Factors}

Other factors for assessing a reactor design include sustainability,
economics, waste management, and non-proliferation.

Breeder \gls{MSR} designs easily score well in the sustainability category.
\glspl{MSR} generally have good neutron economy due to little structural
material in the core and continuous online removal of neutron poisons. Noble
metal neutron poisons also naturally plate out onto the inner walls of the
loop which are generally regions of lower neutronic importance. Both
$^{232}$Th/$^{233}$U and $^{238}$U/$^{239}$Pu fuel cycles are viable
candidates for breeding in \glspl{MSR}, with the former being more suited for
thermal reactors and the latter being more suited for fast reactors. Another
benefit of this is that some \glspl{MSR} can potentially start with an initial
core loading of $^{239}$Pu with a fast spectrum before transitioning towards
$^{233}$U fuel with a $^{232}$Th feed; this measure effectively circumvents
the issue of having insufficient $^{233}$U inventory to start up the reactor.

The $^{232}$Th/$^{233}$U fuel cycle produces significantly less \gls{TRU}
waste than the other cycles due to the smaller atomic masses of $^{232}$Th and
$^{233}$U. This reduces the overall radiotoxicity and long-term decay heat
associated with long-lived plutonium and \gls{MA} isotopes. However, there are
other radionuclides such as $^{231}$Pa, $^{229}$Th, and $^{230}$U that may
pose long-term radiological concerns. Nevertheless, this feature complements
\gls{TRU} burning in fast spectrum \glspl{MSR} to reduce overall levels
of \gls{TRU} waste going into long-term storage in nuclear waste repositories.

There are some nuclear non-proliferation concerns with the thorium fuel cycle
in \glspl{MSR}.
The main concern involves the separation of the intermediate $^{233}$Pa
isotope from the fuel salt. $^{233}$Pa decays into $^{233}$U with a half-life
of approximately 27 days and $^{233}$U is at least as potent as $^{235}$U for
nuclear weapons production. The highly radioactive $^{232}$U provides some
level of proliferation resistance but it can be sidestepped by
separating away its $^{232}$Pa precursor, which has a half-life of only 1.31
days. Strong regulatory framework and close monitoring of \glspl{MSR} would be
essential to ensure non-proliferation.

The economic prospects are uncertain at the current stage of \gls{MSR}
development but we can qualitatively list the potential impacts. Capital costs
are expected to be greater due to the need for increased corrosion resistance
and the supporting reprocessing facilities. This is in spite of the reduced
pressure constraints on the reactor pressure vessel for \glspl{MSR} resulting
from the significantly lower operating pressure. There are potential cost
savings from eliminating fuel pellet and fuel assembly fabrication. Higher
fuel utilization is achievable as neutron poisons are continuously removed
from the core via the continuous online reprocessing and gas sparging systems.
Online fuel reprocessing also minimizes reactor downtime because refuelling
can be performed at any time. The higher operating temperature results in
greater thermal efficiencies and presents new opportunities beyond
conventional electricity generation, namely industrial process heat
applications and hydrogen production.

Beyond these factors, \glspl{MSR} still require significant R\&D efforts and
engineering demonstrations for experimental validation of various components
before a full commercial model can be commissioned. Work towards creating a
safety and licensing framework for \glspl{MSR} has picked up pace only in
recent years due to the growing interest from commercial \gls{MSR} developers.

>>>>

\glspl{MSR} are one of six advanced reactor designs shortlisted by
the Generation IV Forum in 2001 for promising significant advances in safety,
sustainability, efficiency, and cost over existing designs. This
has attracted significant attention and resources towards
\gls{MSR} research, most noticeable by the number of start-up companies that
have emerged in recent years touting various \gls{MSR} designs. This chapter
provides a brief history of \glspl{MSR}, followed by the distinctive features
that earned the concept the label of being a Generation IV reactor. Lastly,
this chapter presents the reference specifications of the \gls{MSFR} concept
studied in this work.

@@

The first \gls{MSR}, named the \gls{ARE}, dates back to the 1940s
as part of the US Aircraft Nuclear Propulsion program
\cite{rosenthal_molten-salt_1970}. Researchers recommended the molten fluoride
salts in particular for high uranium solubility, chemical stability, low vapor
pressure even at high temperatures, decent heat transfer properties,
resistance against radiation damage, and reduced corrosive effects on some
common structural material \cite{rosenthal_molten-salt_1970}. They
subsequently built the 2.5 MW$_{\text{th}}$ ARE reactor at \gls{ORNL}, where
it achieved criticality on November 1954 and generated 100 MWh over nine days.
The fuel consisted of enriched uranium in a molten salt mixture of NaF,
ZrF$_4$, and UF$_4$, and the reactor used blocks of BeO for neutron
moderation. The aircraft program
ultimately never came to fruition as the development of intercontinental
ballistic missiles effectively eliminated the need for long-range
nuclear-powered bomber aircraft.

However, the successful demonstration of the \gls{ARE} spurred further
research into adapting \glspl{MSR} for civilian power generation
\cite{rosenthal_molten-salt_1970}. One key finding from the
research was that the thorium fuel cycle had a better breeding ratio than the
$^{238}$U-to-$^{239}$Pu fuel cycle in thermal-spectrum reactors.
Ultimately, these efforts culminated in the design, construction, and
successful operation of the \gls{MSRE}, a graphite-moderated thermal
\gls{MSR}. The \gls{MSRE} had a
graphite-moderated design with a LiF-BeF$_2$-ZrF$_4$-UF$_4$ fuel salt mixture,
initially rated at 10 MW$_{\text{th}}$ but later restricted to 8
MW$_{\text{th}}$ due to a miscalculation of heat transfer capabilities
\cite{haubenreich_experience_1970}. In January 1969, the \gls{MSRE} became the
first reactor to run on $^{233}$U fuel.

Building on their experience with the \gls{MSRE}, \gls{ORNL} proposed a
new program for the construction and operation of a demonstration reactor
based on the Molten Salt Breeder Reactor (MSBR) concept that they had
developed \cite{macpherson_molten_1985}. The \gls{MSBR} is a thermal-spectrum,
single fluid reactor with fertile $^{232}$Th isotopes mixed directly into the
FLiBe molten salt for $^{233}$U breeding \cite{gehin_liquid_2016}. Like the
\gls{MSRE}, the \gls{MSBR} relies on continuous online reprocessing to add
fertile material and remove fission product neutron poisons. Researchers
estimated the doubling time (the minimum amount of time required to produce
enough fissile material to start up another \gls{MSBR}) to be
approximately 22 years. However, \gls{ORNL} failed to secure funding for the
new program in their two attempts in 1972 and 1974. Nevertheless, from a
technical perspective, two independent
technology evaluation and design studies of the \gls{MSR} had reported
favorably on the promise of the system \cite{macpherson_molten_1985}.

In spite of this setback, research into \glspl{MSR} continued through the late
1970s. In 1980, \gls{ORNL} published a report describing a new \gls{MSR}
concept, called the \gls{DMSR} \cite{gehin_liquid_2016} with denatured
$^{235}$U fuel (i.e. low-enriched uranium). The \gls{ORNL} researchers
developed this design in response to the fuel reprocessing restrictions
introduced by President Ford in 1976. The \gls{DMSR} would operate as a
once-through
converter system without fuel reprocessing. While the fuel largely consists of
19.75 \% \gls{HALEU}, the initial core loading includes thorium to
boost its conversion ratio throughout its lifetime. It has a continuous
online feed consisting of \gls{HALEU} to maintain criticality, and denatured
$^{235}$U to keep uranium enrichment levels below nuclear non-proliferation
policy thresholds. The design also includes a gas sparging system for removing
gaseous fission products, while noble metals plate out onto the walls of
the coolant loop. The older \gls{MSBR} design had a significant drawback; the
extensive neutron damage in the graphite moderator necessitated frequent
replacement (every four years) throughout its operational lifetime. The
\gls{DMSR} avoids this issue by having a lower power density while maintaining
the overall power output of 2250 MW$_{\text{th}}$. As a result, researchers
projected that the graphite moderator would last for the entirety of the
\gls{DMSR}'s design lifetime.

There was a concurrent program at the UK Atomic Energy Authority for the
development of a 2500 MW$_{\text{e}}$ lead-cooled
\gls{MCFR} concept \cite{smith_assessment_1974}. It is a dual fluid system,
with separate loops for the fuel salt and the blanket salt. The blanket is a
1 m-wide tank surrounding the core. The absence of moderators and the choice
of chloride over fluoride salt resulted in a relatively hard neutron spectrum
which favors $^{239}$Pu breeding over the thorium cycle. The UK researchers
performed some experiments to study molten salt chemistry but there were no
reactor prototypes. The UK program eventually shut down just like its
US counterpart partly due to the successful demonstration of the Prototype
Fast Reactor which had achieved criticality in 1974.

Following a lull lasting through the late 20th century, \gls{MSR} research
picked up pace due to renewed interest initiated by the Generation IV Forum in
2001. Today, there are numerous \gls{MSR} concepts under active development
led by various national and commercial bodies.

@@

As mentioned in Chapter \ref{chap:intro}, the most significant difference
between \glspl{MSR} and other reactor concepts is the liquid fuel in
\glspl{MSR}; fissile and/or fertile material is dissolved in high temperature,
commonly eutectic mixtures of molten salts. Molten salt-cooled, solid-fuel
reactors also exist but this thesis will focus on liquid-fuel reactors.
The primary coolant loop containing the fuel salt
transfers heat through a heat exchanger to the clean, intermediate
loop. The liquid fuel form allows for continuous online fuel reprocessing,
and the removal of gaseous fission products via a gas sparging system.

The various \gls{MSR} designs under development today illustrate the
flexibility of this reactor concept. Graphite-moderated thermal-spectrum
\glspl{MSR} are typically straightforward \gls{LEU} burners or
$^{232}$Th/$^{233}$U iso-breeders/breeders, while epithermal- and
fast-spectrum \glspl{MSR} can operate as \gls{TRU} fuel burners or
$^{238}$U/$^{239}$Pu breeders. Breeder designs can be further categorized into
one- or two-fluid designs. Two-fluid designs feature separate blanket molten
salt mixtures that contain higher proportions of fertile material than the
fuel salt mixture. Examples of one-fluid designs include the Integral Molten
Salt Reactor from Terrestrial Energy \cite{leblanc_integral_2015} and the
\gls{MSR} design from Transatomic Power
\cite{transatomic_power_corporation_technical_2016} while two-fluid designs
include the \gls{MSFR} \cite{serp_molten_2014} and \gls{MOSART}
\cite{ignatiev_molten_2014}.

@@

\glspl{MSR} rely on natural physical phenomena for passive safety such as the
strong negative temperature reactivity feedback of the fuel salt due to
greater temperature-induced expansion in liquid fuel than solid
fuel. Combined with the Doppler broadening of resonance capture cross sections
present in both fuel forms, there would be a smaller temperature
increase following an unprotected reactivity insertion. The overall
temperature reactivity coefficient varies widely among \gls{MSR}
designs due to other structures, such as moderators and reflectors, present in
the core. In particular, graphite moderators tend to have slightly positive
temperature reactivity coefficients. \gls{MSBR} concept has this issue, but
the total temperature reactivity coefficient is still relatively large and
negative \cite{rykhlevskii_modeling_2019}. The negative temperature reactivity
feedback provides a large degree of control and stability as it is always
present in an \gls{MSR} regardless of the operating conditions.

Continuous online fuel reprocessing allows operators to maintain low excess
reactivity inventories in the core as additional fuel can be added on an ad
hoc basis. Reprocessing and gas sparging systems help reduce fissile
requirements by continuously removing neutron poisons. These factors, in
addition to the strong negative temperature
reactivity coefficient, diminish the likelihood and severity of unprotected
criticality accidents in \glspl{MSR} \cite{elsheikh_safety_2013}. In the
unlikely situation in which an \gls{MSR} encounters a severe runaway reaction,
\glspl{MSR} rely on another passive safety feature: freeze plugs. Various
freeze plugs designs exist for different \glspl{MSR}. The freeze plug
in the \gls{MSFR} concept is a plug of solidified salt at the bottom of the
core actively cooled by fans or other cooling systems to keep its temperature
just below the freezing point of the salt \cite{aji_experimental_2020}. In
general, when temperatures in the core exceed a certain threshold during a
dangerous transient, the freeze plug melts and the molten salt in the core
drains into a containment tank designed to keep the salt in a subcritical
configuration. This is especially easy to achieve in thermal-spectrum
\glspl{MSR} as the absence of moderators in the containment tank would
automatically drive the multiplication factor down below unity
\cite{elsheikh_safety_2013}.

During pump failure accidents, natural circulation can passively sustain
enough heat transfer to remove decay heat and prevent catastrophic structural
failure. If natural circulation proves insufficient, the aforementioned freeze
plug can drain the salt out of the core. Decay heat in \glspl{MSR} with online
reprocessing is typically lower than that in \glspl{LWR} due to the
continuous removal of fission products. For example, the decay heat in an
\gls{MSFR} after reaching equilibrium salt composition is approximately
3.5\% of full reactor power compared to 6\% in \glspl{LWR}
\cite{brovchenko_design-related_2013}.

\glspl{MSR} also typically have large margin-to-boiling under nominal
operating conditions so that fuel salt boiling does not occur
\cite{elsheikh_safety_2013}. Furthermore, the reactor vessel is consequently
subject to much lower stresses as \glspl{MSR} operate at near-atmospheric
pressure levels. Thus, the probability of pipe ruptures due to high pressure
is low.

However, there are also some safety disadvantages associated with \glspl{MSR}.
Firstly, \glspl{MSR} have smaller fractions of delayed neutron precursors in
the active core region as some of them decay in the external loop regions.
This complicates reactor control and may result in faster transients due to
the decrease in average neutron lifetime. Secondly, hot molten salts are
generally corrosive and the corrosion mechanism is different from conventional
corrosion induced by water and other common agents \cite{yoshioka_7_2017}. The
intense gamma and neutron radiation in the core may also accelerate corrosion
in the structural components. Lastly, overcooling in pipes and heat exchangers
may pose operational challenges as the salt can freeze and restrict flow,
thereby causing a loss of flow accident \cite{ho_molten_2013}.

\subsection{Other Features}

\glspl{MSR} have several positive sustainability features.
\glspl{MSR} have good neutron economy due to the continuous removal of
neutron poisons by the online reprocessing and gas sparging systems
\cite{kamei_recent_2012}. Both
$^{232}$Th/$^{233}$U and $^{238}$U/$^{239}$Pu fuel cycles are viable
candidates for breeding in \glspl{MSR}, with the former being more suited for
thermal reactors and the latter being more suited for fast reactors.

The $^{232}$Th/$^{233}$U fuel cycle produces significantly less \gls{TRU}
waste than the other cycles due to the smaller atomic masses of $^{232}$Th and
$^{233}$U. This reduces the overall radiotoxicity and long-term decay heat
associated with long-lived plutonium and \gls{MA} isotopes. The combination
of \gls{TRU} fuel and $^{232}$Th feed in fast spectrum \glspl{MSR} would help
reduce overall levels of \gls{TRU} waste going into long-term storage in
nuclear waste repositories \cite{merle-lucotte_launching_2011}.

Nuclear non-proliferation concerns in thorium-fueled \glspl{MSR} involve the
separation of the intermediate $^{233}$Pa
isotope from the fuel salt. $^{233}$Pa decays into $^{233}$U with a half-life
of approximately 27 days and the $^{233}$U produced is equivalent in potency
to $^{239}$Pu for nuclear weapons production \cite{grape_10_2017}. The highly
radioactive $^{232}$U
by-product provides some level of proliferation resistance but nuclear
proliferators can sidestep this complication by
separating the $^{232}$U away from the combined $^{232}$Pa/$^{233}$Pa stream
after most of the $^{232}$Pa has decayed into $^{232}$U at a short half-life
of 1.31 days \cite{grape_10_2017}. Safeguards by design and close monitoring
of \glspl{MSR} are therefore essential to avoiding spent nuclear fuel
diversion.

Economic analyses of \glspl{MSR} are preliminary at the current
stage of \gls{MSR} development. Qualitatively, many technical factors favor
\glspl{MSR} over \gls{LWR}. Some of these factors include: smaller reactor
core size due to low operating pressures, higher thermal efficiency and
cheaper air-cooling systems due to high operating temperatures, reduction in
fuel fabrication costs, and shorter reactor shutdown periods due to online
refueling \cite{dolan_1_2017}. Moir \cite{moir_recommendations_2008}
published data from \gls{ORNL} reports on the estimated electricity costs for
a 1000 MWe \gls{MSR} with 20\% enriched $^{235}$U fuel and no reprocessing
throughout the reactor's 30-year lifetime. Adjusted to year 2000 dollars, the
electricity costs for this type of \gls{MSR} was 7\% and 8\% lower than that
of pressurized water reactor and coal power plants. These differences are
within the $\pm$10\% uncertainty range but there is room for some optimism.
The author also notes that costs could be lower in an \gls{MSR} with
fertile feed and online reprocessing could reach fuel self sufficiency.

In terms of technological readiness, \glspl{MSR} still require significant
R\&D efforts and engineering demonstrations for experimental validation of
various components before a full commercial model can be commissioned. Work
towards creating a safety and licensing framework for \glspl{MSR} has picked
up pace only in recent years due to the growing interest from commercial
\gls{MSR} developers.

\section{Literature Review of MSR Simulation Tools}

In the past two decades, researchers have developed several
new tools for simulating steady-state and transient behavior in \glspl{MSR}.
Many of the earlier efforts featured simplifications in simulating
thermal-hydraulics by solving 1-D Navier-Stokes equations or using
predetermined uniform velocity fields \cite{krepel_dyn3d-msr_2007,
kophazi_development_2009}. In more recent years, there has been
significant progress towards fully coupled, spatial codes that feature
2-D axisymmetric or full 3-D models. In 2011, Cammi et al.
\cite{cammi_multi-physics_2011} performed a ``\gls{MPM}'' analysis
of a simplified 2-D axisymmetric model of a single \gls{MSBR} fuel channel
using the commercial finite element analysis software COMSOL Multiphysics. The
physics were implemented through the two-group neutron diffusion equations,
and the \gls{RANS} standard $k$-$\epsilon$ turbulence model, for the
neutronics and thermal-hydraulics respectively. The
authors emphasized the need for proper full coupling of the multiphysics and
presented both steady-state and transient results in various
scenarios such as reactivity insertions, changes in pumping rate, and the
presence of periodic perturbations. This approach was featured again in a
later paper by Fiorina et al. \cite{fiorina_modelling_2014} in 2014 for a 2-D
axisymmetric model of the
\gls{MSFR}. The authors presented results from the Politecnico di Milano
COMSOL-based approach, and another approach by researchers from Delft
University of Technology, in which they coupled their in-house neutronics and
thermal-hydraulics codes, DALTON and HEAT respectively. With multigroup
neutron diffusion and \gls{RANS} formulations on ultra fine meshes, both
models showed good agreement in the steady-state neutron flux, temperature,
and \gls{DNP} distributions, and in the power responses following various
accident transient initiations. Aufiero et al. \cite{aufiero_development_2014}
concurrently developed
a full-core 3-D model of the \gls{MSFR} on OpenFOAM, albeit with one-group
neutron diffusion to reduce computational load. With the 3-D model, the
authors could simulate the asymmetric reactor response to the failure of a
single pump in the sixteen-pump \gls{MSFR} configuration. The authors also
provided some quantitative data supporting the use of implicit coupling over
explicit coupling to obtain accurate solutions of the transient cases.
Recognizing the huge computational burden required for full 3-D simulations, 
later authors came up with innovative ways to alleviate this issue such as
selective geometrical reduced order modeling for various components of a
reactor based on the importance of the physical phenomena being simulated
\cite{zanetti_geometric_2015}, or using a novel, efficient method for
neutronics calculations \cite{laureau_transient_2017}.

\section{Molten Salt Fast Reactor}

The \gls{MSFR} is a European reference fast-spectrum \gls{MSR} concept
\cite{mathieu_thorium_2006, merle_optimized_2007}. Table \ref{table:msfr} and
Figure \ref{fig:msfr} show the main specifications and schematic view of the
\gls{MSFR}, respectively. Developed from the \gls{MSBR} design, the
\gls{MSFR} is intended to run primarily on a closed thorium fuel cycle with
continuous online fuel reprocessing. Several reasons motivated the omission of
graphite moderators from the original \gls{MSBR} design. Graphite is
susceptible to
long-term radiation damage and replacement is likely to be necessary during
the operating lifetime of the reactor. Graphite also has a positive
temperature coefficient of reactivity; eliminating graphite from the design
ensures a greater safety margin \cite{mathieu_thorium_2006}. While negative
temperature coefficients are attainable with very thermalized spectra,
breeding ratios deteriorated significantly due to parasitic absorption in the
large volume of graphite needed for thermalization
\cite{mathieu_thorium_2006}.
%
\begin{figure}[htb!] 
	\centering
	\includegraphics[width=0.8\textwidth]{MSFR}
	\caption{Schematic view of the MSFR concept. Figure reproduced from
	Brovchenko et al. \cite{brovchenko_design-related_2013}.}
	\label{fig:msfr}
\end{figure}
%
\begin{table}[htb!]
    \small
	\caption{Main specifications of the \gls{MSFR} concept
				\cite{serp_molten_2014}.}
	\centering
	\begin{tabular}{ l r }
		\hline
		Parameter & Value \\
		\hline
		Thermal/Electric output [MW$_{\text{th}}$/MW$_{\text{e}}$] & 3000 /
		1500 
		\\
		Salt volume [m$^3$] & 18 \\
		Salt fraction in core & 0.5 \\
		Number of circulation loops & 16 \\
		Nominal flow rate [kg s$^{-1}$] & 18500  \\
		Nominal circulation time [s] & 4.0 \\
		Inlet/outlet temperature [K] & 923 / 1023 \\
		Blanket volume [m$^3$] & 7.3\\
		\hline
	\end{tabular}
	\label{table:msfr}
\end{table}

In the \gls{MSFR} design, fuel salt flows vertically upward through a 9 m$^3$
central core region. At the top of the core, the flow separates into sixteen
smaller external loops, each of which passes through a heat exchanger before
being pumped back into the bottom of the core. The salt also passes through
the online salt reprocessing and gas sparging systems located along the
external loops. A toroidal blanket tank containing fertile salt for breeding
surrounds the core radially. The top and bottom of the core are enclosed
by nickel alloy reflectors. A layer of boron carbide behind the blanket tanks
protects the peripheral equipment from excessive neutron damage. During severe
accidents when core temperatures rise to dangerous levels, the actively
fan-cooled freeze plug at the bottom of the core melts and drains the fuel
salt into a containment vessel designed to keep the salt subcritical.

Although the \gls{MSFR} is primarily designed to operate on the thorium fuel
cycle, it supports a range of start-up fuel and feed compositions. This
versatility is particularly important for the first few \glspl{MSFR} to be
deployed due to the lack of $^{233}$U reserves required for the initial core
loading. In general, the fuel and blanket salts are approximately composed of
eutectic mixtures of 77.5\% LiF - 22.5\% AcF$_4$; AcF$_4$ represents
actinide fluorides such as uranium, thorium, plutonium, and other \gls{TRU}
fluorides. For an initial composition consisting of $^{232}$Th and $^{233}$U,
the benchmark value for the amount of $^{233}$U for criticality under
normal operating conditions is 2.5 mol\%. However, a neutronic benchmark
study by Brovchenko et al. \cite{brovchenko_neutronic_2019} shows that
different neutronics software with different nuclear data clearly provide
different $k_{eff}$ estimates even with the same isotopic compositions and
temperature distributions. The individual co-authors adjusted the
ratio of $^{232}$Th to $^{233}$U slightly to achieve exact criticality at a
uniform temperature of 973 K for their own neutronics software
\cite{brovchenko_neutronic_2019}. This thesis
performs the same exercise to adjust the inlet and outlet temperatures to
match nominal values.

The thermal and electric power output of the \gls{MSFR} are 3000 MW
$_{\text{th}}$ and 1500 MW$_{\text{e}}$, respectively. The high thermal
efficiency ($\eta_{th}=0.5$) is due to the high operating
temperature. The inlet and outlet temperature
specifications of the fuel salt are 923 K and 1023 K, respectively, for a
minimum 50 K temperature buffer between the
operating temperatures and the melting point of the salt (873 K)
\cite{euratom_final_2015}. The \gls{MSFR} has
heat exchangers and an intermediate coolant loop to isolate the power
conversion system from the highly radioactive fuel salt. This also serves as
a layer of containment between the radioactive material and the outside
environment. The exact composition of the intermediate coolant is not
finalized yet but potential candidates include NaF-NaBF$_4$, FLiNaK,
LiF-ZrF$_4$, and FliBe \cite{merle_concept_2017}.

\subsection{Model Reactor Geometry}
